\forestclaw: Hybrid forest-of-octrees AMR for hyperbolic conservation laws

### 2013-10-31 docs/parco13/paper.tex
Tweak section on parallelization

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The \ignore{order of leaves} leaf ordering
is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves
(see \Fig{cubed3}).

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.  The structure of \forestclaw is oblivious to the fact that it only has a
local view of the distributed mesh and data which relieves the developer from
programming to the MPI interface.
% Parallel neighbor exchanges are hidden
% inside \pforest and called by \forestclaw
% % at well-defined synchronization points
% as follows.

With MPI parallelism, neighbor patches can be either local or assigned to a
different process (ghost patches).  Since this must make no difference
numerically, we need to ensure that the values of parallel neighbor patches are current
whenever they are accessed.  To this end, we allocate storage for a layer of
ghost patches in \forestclaw, which we can pass to a general-purpose \pforest
routine that communicates local leaf data to all processes that view a
particular leaf as a ghost.  If we call this routine before we go through the
local neighbor interactions, we can handle ghost values implicitly by neighbor
lookups without querying if they are local or remote.

@@

% count of exchanges with and without sub-cycling
In the context of finite-element or finite-difference methods, there should be
one such parallel data exchange per time step for a global value of the time
step length $\dt$, or one exchange per discretization level per time step if
$\dt$ is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  In the tradition of block-structured AMR codes, interaction
between neighbor patches is done in a hierarchy from coarse to fine levels, and
then correction factors are propagated back from fine to coarse levels, requiring
an exchange at each level.  With sub-cycling, this entails a recursion with an
operation count that is exponential in the difference between the largest and
the smallest refinement level, with the benefit that $\dt$ matches the CFL
condition at each level.  While a large number of exchanges per time step can
present a scaling bottleneck due to the inherent synchronization and latency
losses, the expectation is still that the time to solution improves when
switching from uniform to adaptive meshes, since we are computing with fewer
patches, and improves again when enabling sub-cycling, since we take larger
time steps on the coarser levels.

>>>>

The MPI layer is addressed from within \pforest and not exposed to the
\forestclaw code.  The \ignore{order of leaves} leaf ordering
is maintained in \pforest according to a
space filling curve.  Each MPI rank has a local view on its own partition,
augmented where necessary with information about one layer of ghost leaves
(see \Fig{cubed3}).
The structure of \forestclaw is oblivious to the fact that it only has this
local view, which relieves the developer from programming to the MPI interface.

@@

When \forestclaw accesses neighbor patches, they can be on the same or a
different block.  In the latter case, coordinate transformations are carried
out.
% Parallel neighbor exchanges are hidden
% inside \pforest and called by \forestclaw
% % at well-defined synchronization points
% as follows.
Irrespective of their block,
% With MPI parallelism,
neighbor patches can be either local or assigned to a different MPI process (in
which case we call them ghost patches).  Since this must make no difference
numerically, we need to ensure that the values of parallel neighbor patches are current
whenever they are accessed.  To this end, we allocate storage for a layer of
ghost patches in \forestclaw, which we can pass to a general-purpose \pforest
routine that communicates local leaf data to all processes that view a
particular leaf as a ghost.  If we call this routine before we go through the
local neighbor interactions, we can handle ghost values implicitly by neighbor
lookups without querying if they are local or remote.

@@

% count of exchanges with and without sub-cycling
In the context of finite-element or finite-difference methods, there should be
one such parallel data exchange per time step for a global value of the time
step length $\dt$, or one exchange per discretization level per time step if
$\dt$ is chosen per-patch depending on its size (this is sometimes called
sub-cycling).  In the tradition of block-structured AMR codes, interaction
between neighbor patches is done in a hierarchy from coarse to fine levels, and
then correction factors are propagated back from fine to coarse levels, requiring
at most two exchanges per level.  With sub-cycling, this entails a recursion with an
operation count that is exponential in the difference between the largest and
the smallest refinement level, with the benefit that $\dt$ matches the CFL
condition at each level.  While a large number of exchanges per time step can
present a scaling bottleneck due to the inherent synchronization and latency
losses, the expectation is still that the time to solution improves when
switching from uniform to adaptive meshes, since we are computing with fewer
patches, and improves again when enabling sub-cycling, since we take larger
time steps on the coarser levels.
